---
title: "Introduction to R and dplyr"
author: "Jasmine"
date: '2019-04-03'
output: html_document
---
## To do before hand:
1. install tidyverse and rstudio:
Follow setup instructions for R and Rstudio if you don't have them already: https://datacarpentry.org/R-ecology-lesson/

2. installing packages
- one time thing (you can use library(tidyverse) to check if it is installed -throws an error if it isn't)
```{r}
#uncomment the line below
#install.packages("tidyverse", "readxl")
```

## objectives 
- read in files of different formats
- apply the help() documentation to helping you learn new functions and debugging
- identify resources and tactics in debugging your own code
- using the dplyr library in the tidyverse to manipulate your data in a datafram
- save your file into a different directory

## Acknowledgements
This document draws heavily on (and also really good resources for future reference):

- Jenny Bryan: stats professor @ UBC! now at Rstudio [what they forgot in R](https://whattheyforgot.org/index.html)
- [R for Data Science](https://r4ds.had.co.nz/)
- [Data Carpentry](https://datacarpentry.org/lessons/#ecology-workshop)

***
# Workshop Part 1

## R basics
- Going through them really quickly just so you have an idea of them
- I have integrated them into the Tidyverse part of the workshop to get you used to seeing them in the wild and actually have context for why it is important

### Rstudio - Integrated Development Environment (IDE)
- get familiar with the layout
- R (language) & Rstudio the interface that facillitates your interactions with R
- usually it is advised to change your "save workspace to .RData on exit" to never to increase reproducibility

### rProjects
- easier to keep your files together
- don't need to worry as much with dealing with directries

```{r}
# Try making one now --> top left with the blue cube
```
- it is actually just a file that rstudio knows to identify
- will work the same way as a folder

### Rmarkdown
- what this file is made with!
- a slightly different way to run your r code
- instead of commenting out your text you can write it outside of the code chunk
```{r}
#this is a code chunk

#your r code goes in here
```

use ctrl + alt + i to make a new code chunk here:


```{r}
#Run this cell --> (a) click on the green arrow
2 + 2
#(b) select the code you want to run or be in the line and cmd + shift + enter
```
^ output is right below instead of in the console below

### Assignment
although = works but it is convention to use <- 

```{r }
x <- 2 + 2
# you can say it in your head as: x gets the result of 2 + 2
```

__pro tip__: use `alt + -` to quickly make the assignment symbol and the spaces needed!

- To your right is the `Environment` which shows you what is in `memory` containing the variables that R is currently holding for you that you can ask for
- bellow is the `console` although using Rmarkdown reduces the need for it but you can still see the lines of code was sent there to be run (think of the Rmd file as your scrap paper)

***

If you ever need a refresher...
 
#### shortcuts cheatsheet (on Mac)
 __replace cmd with ctrl for windows in most cases__
 1. run a cell `shift + enter`
 2. assignment <- `alt + -`
 3. multi-line select `ctrl + option + arrow key`
 4. pipe %>% `cmd + shift + m`

### Types
#### Objectives
- figuring out what type an variable is using the function `str()`

Often over looked but pretty important... a major cause of errors
- will be highlighted in Rstudio (good way to double check)
- you can also check if something has a certain type by using the ```str()``` (structure) function
- functions are a command that tells R to excute a set of tasks previously defined
- functions are structured like this: name_of_fn(argument/parameter)

```{r types}
str(13) # numeric
str("I am a string") # character
str(TRUE) # logical
```
* note because I came from Python I may accidently use the other terms

You can also change the types by using ```as.[insert type here]```
- you can use `tab` completion to 
```{r changing types}
one <- ("1")
str(one)

one <- as.numeric(one)
str(one)
```

#### Practice
try our changing the type numeric -> character
```{r}

```

keep all of this in mind we will be seeing this in the next part

### Opening Files
#### objectives
- opening different file types
- directories

***

#### CSV - comma seperated files
- most functions that imports files into R start with `read`
- use that to our advantage when using tab completion even when we don't know what the function might be called exactly
```{r}
#tab completion demo
```

not sure how to use it? call for help!
- name_of_function{what package it belongs to}
- description
- arguments
- scroll to the bottom for examples
```{r}
help("read.csv")
# look in the help pane on the lower left
```

now use the function:
```{r results="hide"}
#this will work
#df <- read.csv("data/PTM-Master_List.csv")
```

#### Practice
apply what we learned in types to check to see what type is df
```{r}

```
Congrats! It is a dataframe! Another type of data (like character, logical etc.) and each column has a type

#### Directories
```{r}
#this will not work
#read.csv("PTM-Master_List.csv")
```

__note__: 
*sketch this out*
directories - "folder"
- important to know where your "folder" is

this is the path - the location of your folder:
```{r}
getwd()
```
your files should be in the location specified

__Pro tip__: use Rprojects to help you get organized and you can worry less about directories and easier to share your work
- although you can use `setwd()` but it is generally not recommended because other people may have a different file structure than you
- there are packages like `here()` that make it easy for you to get your path

```{r results = "hide"}
df <- read.csv("data/PTM-Master_List.csv")
df <- read.csv("./data/PTM-Master_List.csv")

#notice what is in your console - parsed with column specification
#auto detects your types

#notice that this does not work because it does not include the entire path name
#df <- read.csv("/data/PTM-Master_List.csv")

#the . you can think of it as everything before it
#like this
full_path <- paste0(getwd(),"/data/PTM-Master_List.csv")
```

### read_csv (tidyverse)
- Tidyverse also includes a read csv function
- they can also be loaded seperately
```{r results='hide'}
library(tidyverse)
```

```{r results ="hide"}
df_1 <- read_csv("data/PTM-Master_List.csv")
df_2 <- read_csv("./data/PTM-Master_List.csv")
```

***
### Practice
Apply what you learned to a similar function that works with excel files
- you will need the library readxl
```{r}
# note - there is a file .xlsx in the data folder for you already



#load your package from library
#error? remember to install.packages()



#use tab completion to your advantage to find what function will work




#need help with that function?



#use the function and store it as df_xlsx



```

#### googlesheets
- for people who feel a bit adventurous
```{r results='hide'}
library(googledrive)
#googlesheets::gs_read()

library(googlesheets)
#library(googlesheets4) # future will be googlesheets4
#(ptm <- drive_get("PTM - Master List"))

#need to authenticate
mysheets <- gs_ls()

#gets the sheet of that name
gs_master <- gs_title("PTM - Master List")

#gets the data
gs_ptm<- gs_read(gs_master)

#cleans up columns
colnames(gs_ptm) <- gs_ptm[1,]

#removes the first two rows
new <- gs_ptm[-c(1,2),]
```

### navigating your dataframe
Get a preview of what is inside by using `head()` and `tail()` to see the first few and last few rows
```{r results='hide'}
head(df)

tail(df)
```

Want to know the names of your columns?
```{r}
names(df)
```

Need to refer to a particular row? `DataFrameName$ColumnName`
__Pro tip:__ use tab complete to find column names after your type the $
```{r results="hide"}
df$PTM
```
__Pro tip:__ weird column names with spaces? use '` `' around the Column name

#### Practice
- try doing this with the df_xlsx dataframe
```{r}

```

***
#### BRAIN BREAK
***
### Workshop part 2

### dplyr package
mainly used for cleaning up your data and manipulating your data frame
- how to get certain rows you want using `filter()`
- how to get summary statistics using `group_by()` and `summarise()`
- how to get columns you want `select()`

#### filter()

##### `==` equals

Quick primer on evaluating logicals
```{r}
3 >= 4
3 > 3

"Red" == "Red"

"Red" == "red"
```

This applies to when we want to filter something out of the dataframe
- if it evaluates to `TRUE` then the row will be kept
```{r}
test <- filter(df, df$Red.Green.Brown == "Red")
```

##### booleans 
- ! (not)
- & and
- | or
```{r results="hide"}
#Filter only keeps rows that evaluate to TRUE

#not 
"Red" != "Red"
"Red" != "Blue"
filter(df, df$`Red.Green.Brown` != "Red")

# or
TRUE | TRUE
TRUE | FALSE
FALSE | FALSE
filter(df, df$`Red.Green.Brown` == "Red" | df$`Red.Green.Brown` == "brown")
# TRUE  in one of the cases

# and
TRUE  & FALSE
TRUE  & TRUE
FALSE  & FALSE

filter(df, df$`Red.Green.Brown` == "Red" & df$Extracted. == "x")
# TRUE in all the cases

#sometimes error messages are helpful
#filter(df, df$`Red.Green.Brown` = "Red")
```

tying it together
- this does not work as expected
```{r results="hide"}
filter(df,(df$PTM. < 500))
#read the error message
```

let's check the type of the `df$PTM.`
```{r results="hide"}
str(df$PTM.)
```

we can fix this by changing it to a factor
```{r results="hide"}
df$PTM. <- as.numeric(df$PTM.)

#check it
str(df$PTM.)
```

#### Practice
try filtering for all columns that are not brown algae and between PTM 34 - PTM 400
```{r}

```

##### removing NA
- NA is a data type but slightly different than the rest
- data frames that are read in automatically converts empty cells to NA in read_csv
```{r}
#you can use this to remove NA cells
#filter(df, !is.na(df_1$`PTM.`))

#note # is a special character like $ %
#make sure you use the ``
```

##### `%in%`
a special character
```{r results="hide"}
want <- c(1,13,14,20)

filter(df, df$`PTM.` %in% want)

help("%in%")
```

#### group_by() & summarize()
- usually used together because on it's own isn't terribly useful
- we can use it to get the number of rows using `n()`
```{r}
summarise(df, number = n())
```

__note__: becareful if you have plyr (another library) - reading what is outputted is helpful
```{r}
#library(plyr)
# you can use the :: notation to specify which one you want
#dplyr::summarise(df, number = n())
```

#### introduction to pipes (less naming!)
- think of ` %>% ` as then if you are reading it aloud
- you can use the pipe to "pass one thing that you are using along

```{r results="hide"}
#without pipe
grouped_df <- group_by(df, df$`Red.Green.Brown`)
summary_df <- summarise(grouped_df, number = n())

#with pipe
df %>% 
  group_by(df$`Red.Green.Brown`) %>% 
  summarise(number = n())
```

you can do the same with `count() mean(), max(), min()` to get summary statistics

#### select()
the columns you want
```{r results="hide"}
df %>% 
  select(`PTM.`)
```

#### mutate()
   -> Debugging! - types (boolean, character , syntax highlighting! etc.)
   - print or get the traceback
   - sometimes your function is masked (has the same name)
```{r}
new_df <- df %>% 
  mutate(collector_no = paste("PTM",`PTM.`))
# a column can be given the same name but will be overwritten

#look at the new resulting columns
head(new_df$collector_no)
```

#### Practice
1. keep only the rows with ther Other.collectors with `Liam Coleman`
2. combine the main collector and the other collectors column to a new column called `all_collectors`
```{r}

```

#### Looking for Help (just in case I forget to cover it)
1. Read the error message and go to the line
2. Google is your best friend
  - Try typing in the error message
  - look into the help documentation to see what examples are given and what arguments it takes

***
#### Extra Credit
- gather() spread() if there is time
- join multiple dataframes - `left_join()` , `cbind()`, `rbind()`

#### Saving your file (xlsx, csv)
look at the File tab to see where it is saved
```{r}
write.csv(new_df,file = "new_file.csv")

#get it into the folder
write.csv(new_df,file = "results/new_file.csv")
```

#### Extra Credit
```{r}
library(openxlsx)

write.xlsx(new_df[1:5,1:5], file = "excel_file.xlsx")
```